---
title: "Expectation-Maximization Algorithm: A Method to Solve Latent Variable Problems and Other Latent Variable Problems" 
draft: false
summary: 'This is a summary of the post that brielfy explains the main points of the post to provide an anchor for the reader.' 
article_type: technical
output:
  bookdown::html_document2:
     keep_md: true
always_allow_html: true
bibFile: content/technical_content/em_algorithm/biblio.json    
tags: []
---   


```{r package_loading_1, include=F}
#load packages   
library(easypackages) 
packages <- c('devtools','tidyverse', 'RColorBrewer', 'parallel', 'data.table', 'kableExtra', 'ggtext', 'egg', 'papaja', 'shiny',
              'ggbrace', 'cowplot', 'knitr', 'RefManageR', 'reticulate', 'gluedown', 'formatR')
libraries(packages)  

#use_python(python = "/usr/local/bin/python3.9")

knitr::opts_chunk$set(comment = NA)
# knitr hook to use Hugo highlighting options
knitr::knit_hooks$set(
  source = function(x, options) {
  hlopts <- options$hlopts
    paste0(
      "```", "r ",
      if (!is.null(hlopts)) {
      paste0("{",
        glue::glue_collapse(
          glue::glue('{names(hlopts)}={hlopts}'),
          sep = ","
        ), "}"
        )
      },
      "\n", glue::glue_collapse(x, sep = "\n"), "\n```\n"
    )
  }
)

chunk_class <- function(before, options, envir) {
    class_name = options$class_name

    
    if (!is.null(before)) { 
      
        lines <- unlist(strsplit(x = before, split = "\n")) #separate lines of code at \n
        n <- length(lines)  #determines numbers of lines
        
        #if (line_numbers) { 
           res <- paste0("<pre><code class='", class_name, "'>", before, "</code></pre>")
                            
                            #paste0("<pre><code class='", class_name, "'>", before, "</code></pre>")
        #}
        
       
          
          #res <- paste0("<pre>", paste0("<span class='line-number'>", 1:n,
                            #"</span><code class ='", class_name, "'>", lines, "</code>"), "</pre>")
    }
        return(res)
    
}

knitr::knit_hooks$set(output = chunk_class, preserve = TRUE)

#knitr::knit_hooks$set(output = function(x, options) { 
#  paste(c("<pre><code class = 'r-code'>",
#        gsub('^## Error', '**Error**', x),
#        '</pre></code>'), collapse = '\n')
#})

```


That's some text with a footnote.[^1]$^{,}$ 
[^1]: And that's the footnote (see Table \ref{tab:parameterValues}).

R code chunk below (see lines <a href="#1">1--22</a>). 

```{r r-code, echo=T, eval=F, hlopts=list(hl_lines= paste0('[', 1,',', "2", ",", "5", ']'))}
#this is a comment  more comment my website m , . , y website my website my website
#my website my website my website my website my website my website my website my   
#my website my website my website
print('my website my website my, , , ,. `  website my website my website my website my website
      website my website my   website my website my website my website my website my 
      website')  

print(1 + 2)
mean(x = c(1, 2))
print('another')
print(NULL)
print(NA)
print(TRUE)
"\n"
list('number_measurements' = c(5, 7, 9, 11),
     'spacing' = c('equal', 'time_inc'))

var <- function(x1, x2){

  if (x1 > 2) {print(x1)} 
  else {print (x2)}
}


``` 
This is inline R code:{{< inline-src r >}}print(NULL){{< /inline-src >}}.

Python code chunk below (see lines <a href="#23">23--31</a>). 
```{python, echo=T,  eval=F, hlopts=list(hl_lines= paste0('[', 1,',',3, ',', 4,']'), language = 'python'), engine.path = '/usr/bin/python3', tidy=F}
tup = ['Groucho', 'Marx', 'Xavier', 'Xavier', 'Xavier', 'Xavier', 'Xavier', 'Xavier', 'Xavier', 'Xavier']
list_ex = list(["food", 538, True, 1.454, "food", 538, True, 1.454, "food", 538, True, 1.454, "food", 538, True, 1.454])
sorted(tup)

list_ex2 = list([1 + 2, "a" * 5, 3])  

#deleting variables 
del(list_ex2)
list_ex.count(2)  
```
This is inline Python code: {{< inline-src python >}}print('NULL'){{< /inline-src >}}.

SQL code chunk below (see lines <a href="#32">32--43</a>). 
```{r echo=T, engine = 'sql', eval=F, hlopts=list(hl_lines= paste0('[', 1,',',3, ',', 4,']'), language = 'sql'), tidy=F}
CREATE TABLE person 
  (person_id SMALLINT UNSIGNED,
  fname VARCHAR(20),
  lname VARCHAR(20),
  eye_color ENUM('BR','BL','GR'),
  birth_date DATE,
  street VARCHAR(30),
  city VARCHAR(20),
  state VARCHAR(20),
  country VARCHAR(20),
  postal_code VARCHAR(20),
CONSTRAINT pk_person PRIMARY KEY (person_id)
```
This is inline SQL code: {{< inline-src sql >}} CREATE TABLE person {{< /inline-src >}}.


Javascript code chunk below (see lines <a href="#44">44--59</a>). 
```{r java-code, engine = 'js', echo=T, eval=F, hlopts=list(hl_lines= paste0('[', 2,',',3, ',', 4,']'), language = 'java')}
let codeTable = document.createElement("table");
codeTable.setAttribute('id', "codeTable");

//add rows to table by adding each element of lines
for (let t = 0; t < lines.length; t++) {
  let row = codeTable.insertRow(-1);

  let newCell1 = row.insertCell(0); //insert line number
  let newCell2 = row.insertCell(1);
  let newCell3 = row.insertCell(2);

  newCell1.innerHTML = "<span class= 'line-number' data-number='" + (t+1)  + "'" + "id = '" + 
    (t+1) + "'></span>";
  newCell2.innerHTML = lines[t];
  newCell3.innerHTML = "";
}
```
This is inline Javascript code: {{< inline-src js >}}let codeTable = document.createElement("table");{{< /inline-src >}}. 

CSS code chunk below (see lines <a href="#60">60--65</a>). 
```{r css-code, engine =  'css', echo=T, eval=F, hlopts=list(hl_lines= paste0('[', 2,',',3, ',', 4,']'), language = 'css')}
div[language ='java'] code[data-lang='r'] table td:nth-child(2) { width: 85%;position: relative;

  background-color:  rgba(255,105,130, 0.20);
  border-left: 2pt solid rgba(255,105,130, 0.50);
  padding: 0;
}
```
This is inline CSS code: {{< inline-src css >}} background-color:  rgba(255,105,130, 0.20);{{< /inline-src >}}.

HTML code chunk below (see lines <a href="#66">66--68</a>). 

```{r html-code, echo=T, eval=F, hlopts=list(hl_lines= paste0('[', 2,',',3,']'),language = 'html')}
<script src="{{ "js/external_links.js" | relURL }}"></script>
<script src="{{ "js/number_tables.js" | relURL }}"></script>
<script src="{{ "js/number_figures.js" | relURL }}"></script>
```
This is inline HTML code: {{< inline-src html >}} <script src="{{ "js/external_links.js" | relURL }}"></script>{{< /inline-src >}}.

Bash code chunk below (see lines <a href="#57">69--71</a>). 

```{bash, engine = 'bash',  eval=F, hlopts=list(hl_lines= paste0('[', 1,']'), language = 'bash')}
ls
 
cd ~/Desktop/Home/blog_posts
```
This is inline bash code: {{< inline-src bash >}}cd ~/Desktop/Home/blog_posts{{< /inline-src >}}.


```{r show-results,  class_name = 'r-code', results = 'hold', echo=F, eval=T, hlopts=list(hl_lines= paste0('[', 2,',', "3-5", ']'))}
#this is a comment  more comment my website m , . , y website my website my website
#my website my website my website my website my website my website my website 
#my website my website my website my website my website
print('my website my website my, , , ,. `  website my website my website my website my website my website my website my website my website my website my website my website my website')  
print(1 + 2)
mean(x = c(1, 2))
print('another')
print(NULL)
print(NA)
print(TRUE)
"\n"
list('number_measurements' = c(5, 7, 9, 11),'spacing' = c('equal', 'time_inc'))

var <- function(x1, x2){
shiny::HTML("&nbsp;")

  if (x1 > 2) {print(x1)} 
  else {print (x2)}
}

``` 


Some explanatory text Some explanatory tex {{< cite "fine2019;george2000" >}} Some explanatory text Some explanatory textSome explanatory text{{< cite "fine2019;cole2003" >}}
Some explanatory text Some explanatory textSome explanatory  text{{< cite "fine2019;liu2022" >}}
[link](https://github.com/gohugoio/hugo/issues/9442)

If you liked how these _"generics"_ work in SystemVerilog and how the looks, check out the


```{python, echo=F, eval=T,  class_name = 'python-code', results = 'hold', hlopts=list(hl_lines= paste0('[', 2,',',3, ',', 4,']'), language = 'python'), engine.path = '/usr/bin/python3'}

tup = ['Groucho', 'Marx', 'Xavier']
list_ex = list(["food", 538, True, 1.454])
sorted(tup)

list_ex2 = list([1 + 2, "a" * 5, 3])  

#deleting variables 
del(list_ex2)
list_ex.count(2)  
```


```{r echo=F}
fit = lm(dist ~ speed, data = cars)
b = coef(summary(fit))
```


The slope of the regression is `r b[2, 1]`. This is gamma $\gamma\$. $\frac{1}{2}$ This is (see Equation \ref{eq:multiline}; another comment) 

$$
\begin{alignat}{2}
I & = \int \rho R^{2} dV & + P \nonumber \\\\
Y & = 1 + x
\label{eq:multiline}
\end{alignat}
$$

### Tables

see [section](#section)

 Table \ref{tab:parameterValues}  Table \ref{tab:parameterValues}
Another paragraph begins and the spacing should not be too small from table above. 

 Table \ref{tab:parameterValues1}
 Table \ref{tab:parameterValues}
 Table \ref{tab:parameterValues1}
 
```{r parameterValues, echo=F, eval=T}

#specify parameters for parameter table 
theta <- 3
alpha <- 3 + .32*1
beta <- 180
gamma <- 20

sd_alpha <- 0.05
sd_theta <- 0.05
sd_beta <- 10
sd_gamma <- 4

sd_error <- 0.05


#table of parameter values
parameterValues_df <- data.frame('Parameter Means' = c(
                                         'Baseline, $\\theta$',
                                         'Maximal elevation, $\\alpha$', 
                                         'Days-to-halfway elevation, $\\upbeta$', 
                                         'Triquarter-halfway delta, $\\upgamma$', 
                                         
                         'Variability and Covariability Parameters (in Standard Deviations)', 
                              'Baseline standard deviation, $\\uppsi_{\\uptheta}$',
                              'Maximal elevation standard deviation, $\\uppsi_{\\upalpha}$', 
                              'Days-to-halfway elevation standard deviation, $\\uppsi_{\\upbeta}$',
                              'Triquarter-halfway delta standard deviation, $\\uppsi_{\\upgamma}$',
                         
                              'Baseline-maximal elevation covariability, $\\uppsi_{\\uptheta\\upalpha}$',
                              'Baseline-days-to-halfway elevation covariability, $\\uppsi_{\\uptheta\\upbeta}$',
                              'Baseline-triquarter-halfway delta covariability, $\\uppsi_{\\uptheta\\upgamma}$',
                         
                              'Maximal elevation-days-to-halfway elevation covariability, $\\uppsi_{\\upalpha\\upbeta}$',
                              'Maximal elevation-triquarter-halfway delta covariability, $\\uppsi_{\\upalpha\\upgamma}$',
                         
                              'Days-to-halfway elevation-triquarter-halfway delta covariability, $\\uppsi_{\\upbeta\\upgamma}$',
                          
                              'Residual standard deviation, $\\uppsi_{\\upepsilon}$'), 
                         'Value' = c( theta, alpha, beta, gamma, 
                                     '', sd_theta, sd_alpha, sd_beta, sd_gamma, 
                                     0, 0, 0, 0, 0,0,  sd_error), check.names = F)

#round numbers to that they print with two significant numbers
parameterValues_df$Value <- round(as.numeric(as.character(parameterValues_df$Value)), 3)
parameterValues_df$Value <- formatC(round(parameterValues_df$Value, 3), format='f', digits=2)

#replace '  NA' with empty string 
parameterValues_df$Value[parameterValues_df$Value ==" NA"] <- ''


kbl(parameterValues_df, booktabs = TRUE, format = 'html', 
    align = c('l', 'c'),
    caption = 'Values Used for Multilevel Logistic Function Parameters',    
    escape = F) %>%
    add_indent(positions = c(1:4, 6:16), level_of_indent = 2) %>%
  kable_styling(position = 'center') %>%
  footnote(general =  "<em>Note</em>. The difference between $\\alpha$ and $\\theta$ corresponds to the 50$\\mathrm{^{th}}$ percentile Cohen's $d$ value of 0.32 in organizational psychology (Bosco et al., 2015). Additional text to see what happens",  threeparttable = T,  escape = F, general_title = ' ') %>% 
  row_spec(row = 4, extra_css = 'padding-bottom: 1rem; border-bottom: 1.5px solid') %>%
 row_spec(row = 5, extra_css = 'font-weight: bold; border-bottom: 1.5px solid')
```

```{r parameterValues1, echo=F, eval=T}

#specify parameters for parameter table 
theta <- 3
alpha <- 3 + .32*1
beta <- 180
gamma <- 20

sd_alpha <- 0.05
sd_theta <- 0.05
sd_beta <- 10
sd_gamma <- 4

sd_error <- 0.05


#table of parameter values
parameterValues_df <- data.frame('Parameter Means' = c(
                                         'Baseline, $\\theta$',
                                         'Maximal elevation, $\\alpha$', 
                                         'Days-to-halfway elevation, $\\upbeta$', 
                                         'Triquarter-halfway delta, $\\upgamma$', 
                                         
                         'Variability and Covariability Parameters (in Standard Deviations)', 
                              'Baseline standard deviation, $\\uppsi_{\\uptheta}$',
                              'Maximal elevation standard deviation, $\\uppsi_{\\upalpha}$', 
                              'Days-to-halfway elevation standard deviation, $\\uppsi_{\\upbeta}$',
                              'Triquarter-halfway delta standard deviation, $\\uppsi_{\\upgamma}$',
                         
                              'Baseline-maximal elevation covariability, $\\uppsi_{\\uptheta\\upalpha}$',
                              'Baseline-days-to-halfway elevation covariability, $\\uppsi_{\\uptheta\\upbeta}$',
                              'Baseline-triquarter-halfway delta covariability, $\\uppsi_{\\uptheta\\upgamma}$',
                         
                              'Maximal elevation-days-to-halfway elevation covariability, $\\uppsi_{\\upalpha\\upbeta}$',
                              'Maximal elevation-triquarter-halfway delta covariability, $\\uppsi_{\\upalpha\\upgamma}$',
                         
                              'Days-to-halfway elevation-triquarter-halfway delta covariability, $\\uppsi_{\\upbeta\\upgamma}$',
                          
                              'Residual standard deviation, $\\uppsi_{\\upepsilon}$'), 
                         'Value' = c( theta, alpha, beta, gamma, 
                                     '', sd_theta, sd_alpha, sd_beta, sd_gamma, 
                                     0, 0, 0, 0, 0,0,  sd_error), check.names = F)

#round numbers to that they print with two significant numbers
parameterValues_df$Value <- round(as.numeric(as.character(parameterValues_df$Value)), 3)
parameterValues_df$Value <- formatC(round(parameterValues_df$Value, 3), format='f', digits=2)

#replace '  NA' with empty string 
parameterValues_df$Value[parameterValues_df$Value ==" NA"] <- ''


kbl(parameterValues_df, booktabs = TRUE, format = 'html', 
    align = c('l', 'c'),
    caption = 'Values Used for Multilevel Logistic Function Parameters (see Table \\ref{tab:parameterValues})',    
    escape = F) %>%
    add_indent(positions = c(1:4, 6:16), level_of_indent = 2) %>%
  kable_styling(position = 'center') %>%
  footnote(general =  "<em>Note</em>. The difference between $\\alpha$ and $\\theta$ corresponds to the 50$\\mathrm{^{th}}$ percentile Cohen's $d$ value of 0.32 in organizational psychology (Bosco et al., 2015). see Table \\ref{tab:parameterValues} and Figure \\ref{fig:gg-oz-plot1}.",  threeparttable = T,  escape = F, general_title = ' ') %>% 
  row_spec(row = 4, extra_css = 'padding-bottom: 1rem; border-bottom: 1.5px solid') %>%
 row_spec(row = 5, extra_css = 'font-weight: bold; border-bottom: 1.5px solid')
```


### Figures

```{r gg-oz-plot, eval=F, echo=F}
generate_apa7_figure <- function(fig_label, fig_title, fig_path, fig_note=NA){
  
    html_figure <- paste('<div class= "figure"> <caption class = "figTop"> <span class="figLabel">Figure ', fig_label,  
                         '</span> <br> <span class = "figTitle"> <em>', fig_title, 
                         '</em></span> </caption> <img src=', fig_path, '/>', sep = '')

  if(!is.na(fig_note)) { 
    html_fig_note <- paste('<span class="figNote"><em>Note. </em>', fig_note, '</span> </div>', sep = '')
  }
    
  else{ 
     html_fig_note <- paste('</div>', sep = '')
    }
  
    
  html_complete_figure <- paste(html_figure, html_fig_note)
  
  return(noquote(html_complete_figure))
}

x <- generate_apa7_figure(fig_label = '\ref{fig:gg-oz-plot}', fig_title = 'Title of Figure', fig_path = 'index_files/figure-html/gg-oz-plot-1.png', fig_note = 'Life expectancy from 1952 - 2007 for Australia. Life expentancy increases steadily except from 1962 to 1969. We can safely say that our life expectancy is higher than it has ever been!')


```

<div class= "figure"> <caption class = "figTop"> <span class="figLabel">Figure \ref{fig:gg-oz-plot}</span> <br> <span class = "figTitle"> <em>Title of Figure</em></span> </caption> <img src=images/gg-oz-plot-1.png> <span class="figNote"><em>Note. </em>Life expectancy from 1952 - 2007 for Australia. Life expentancy increases steadily except from 1962 to 1969. We can safely say that our life expectancy is higher than it has ever been!</span> </div>   
 


Figure \ref{fig:gg-oz-plot} 


### References

{{< bibliography cited >}}






